---
title: "Wrangle"
output: html_notebook
---

# Introduction 

Data wrangling: getting the data in a useful form. 
Parts: import, tidy, transform.

# Tibbles

Tibbles are data frames, but tweaked. 

Old data frames can be converted with 'as_tibble()'.

```{r}
library(tidyverse)
as_tibble(iris)
```

To create a new one, use 'tibble()'.

```{r}
tibble(
    x = 1:5,
    y = 1,
    z = x ^2 + y
)
```

It's possible to have column names that are nonsyntatic, i.e., not variable names.

```{r}
tb <- tibble(
    `:)` = "smile",
    ` ` = "space",
    `2000` = "number"
)
tb
```

Another way to create is using transposed tibble 'tribble()'.

```{r}
tribble(
    ~x, ~y, ~z,
    #--|--|----
    "a", 2, 3.6,
    "b", 1, 8.5
    )
```

How does Tibble and data.frame compare? 

1. Printing: tibbles only show the first ten rows and the columns that fit on the screen.
2. Subsetting: tibbles never do partial matching. 

To view the entire data set, run 'View()'.

```{r}
import(nycflights13)
nycflights13::flights %>% 
    View()
```

To extract a single variable we use '$' or '[[.]]'.

```{r}
df <- tibble(
    x = runif(5),
    y = rnorm(5)
)
# Extract by name
df$x

df[["x"]]

# Extract by position
df[[1]]
```

To use in a pipe, use a placeholder '.':

```{r}
df %>% .$x

df %>% .[["x"]]
```

To turn a tibble back into a data.frame, we 'as.data.frame()'. This is useful when dealing with older functions. 

```{r}
class(as.data.frame(tb))
```

# Data Import with readr

```{r}
heights <- read_csv("data/heights.csv")
```

Imports the first line as column names.

To skip a n lines, we may use the parameter 'skip = n'. Or use 'comment = "#"' to drop all lines that start with #.

If the data doesn't have column names, pass the parameter 'col_names = FALSE'.

To pass column names, we use 'col_names = c("x", "y", "z")'.

To deal with missing values, we may use, for example, 'na = "."'.

The base R offers 'read.csv()', however it is slower, don't produce tibbles and are less reproducible.

## Parsing a Vector

Parse functions, denoted by 'parse_*()', take a character vector and return a more specialized vector (like a logical, integer, or date).

```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
str(parse_integer(c("1", "2", "3")))
str(parse_date(c("2010-01-01", "1979-10-14")))
```

### Numbers

Since numbers are written differently around the world, (like '.' and ',' to separate the integer and fractional parts of a real number), we can use 'locale' with 'parse_double()' to treat this.

```{r}
parse_double("1,23", locale = locale(decimal_mark = ","))
```

To deal with symbols such as "%" and "$", we can use 'parse_number()'.

```{r}
parse_number("$100")
parse_number("20%")
parse_number("It cost $123.45")
```

To remove grouping characters, we also use 'parse_number()'.

```{r}
parse_number("$123,456,789")

parse_number(
    "123.456.789",
    locale = locale(grouping_mark = ".")
)

parse_number(
    "123'456'789",
    locale = locale(grouping_mark = "'")
)
```

### Strings

readr uses UTF-8. To deal with other encondings, we use 'parse_character()'.

```{r}
x1 <- "El Ni\xf1o was particularly bad this year"
x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"
parse_character(x1, locale = locale(encoding = "Latin1"))
parse_character(x2, locale = locale(encoding = "Shift-JIS"))
```

We can use 'guess_enconding()' to help figure out. 

```{r}
guess_encoding(charToRaw(x1))
guess_encoding(charToRaw(x2))
```

### Factors

To represent categorical values with a known set of possible values, we use factors.

```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "bananana"), levels = fruit)
```

### Date-Times

1. Date: number of days since 1970-01-01
2. Date-time: number of seconds since 1970-01-01
3. Time: number of seconds since midnight

Uses ISO8601.

```{r}
parse_datetime("2010-10-01T2010")
parse_datetime("20101010")
parse_date("2010-10-01")
```

```{r}
library(hms)
parse_time("01:10 am")
parse_time("20:10:01")
```

If the defaults don't work, make the format explicit (see p. 135-136).

```{r}
parse_date("01/02/15", "%m/%d/%y")
parse_date("01/02/15", "%d/%m/%y")
parse_date("01/02/15", "%y/%m/%d")
parse_date("1 janvier 2015", "%d %B %Y", locale = locale("fr"))
```

## Parsing a File

readr tries to figure out the type of each column using 'guess_parser()', but only by reading the first thousand rows. 

A solution: work column by column.

```{r}
challenge <- read_csv(readr_example("challenge.csv"))
problems(challenge)
```

First copy and paste the column specification:

```{r}
challenge <- read_csv(
    readr_example("challenge.csv"),
    col_types = cols(
        x = col_integer(),
        y = col_character()
    )
)
```

Then tweak the type of the column x:

```{r}
challenge <- read_csv(
    readr_example("challenge.csv"),
    col_types = cols(
        x = col_double(),
        y = col_character()
    )
)
tail(challenge)
```

Fixing the y column:

```{r}
challenge <- read_csv(
    readr_example("challenge.csv"),
    col_types = cols(
        x = col_double(),
        y = col_date()
    )
)
tail(challenge)
```

Every 'parse_x()' function has a corresponding 'col_x()' function.

An alternative is to change the amount of rows used for the guessing. 

```{r}
challenge2 <- read_csv(
                readr_example("challenge.csv"),
                guess_max = 1001
            )
challenge2
```

Sometimes, it's easier to diagnose problems reading all columns as character vectors.

```{r}
challenge2 <- read_csv(readr_example("challenge.csv"),
    col_types = cols(.default = col_character())
)
```

If the file is largy, you might set 'n_max' to 10000 or 100000.

If problems persist, read it as a character vector with 'read_lines()' and use string parsing skills.

## Writing to a File

Two functions: 'write_csv()' and 'write_tsv()'. To export a CSV to Excel, use 'write_excel_csv()'.

Always use UTF-8 and ISO8601.

```{r}
write_csv(challenge, "challenge.csv")
```

The type information is lost when you save to CSV.

Alternatives: 
- Use 'write_rds()' and 'read_rds()'. R exclusive.
- Use the *feather* package with 'write_feather()' and 'read_feather()'. Is usable outside of R.

# Tidy Data with tidyr

```{r}

```
```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```